---
title: "cyclistic_data_202206-202306"
author: "Jonnie"
date: "2023-07-31"
output: html_document
---

# Introduction

This R Markdown document presents an in-depth analsis of Cyclistic bike share data from June 2022 to June 2023. The analysis focuses on three key aspects:

1. Ride Length, Count, and distance by Month
2. Ride Length, Count, and distance by Weekday
3. Preferred Ride Type by Membership

## Data Loading

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)
library(lubridate)
library(geosphere)
library(dplyr)
library(scales)

cyclistic_year_date <- rbind(read.csv("202206-divvy-tripdata.csv"),
                             read.csv("202207-divvy-tripdata.csv"),
                             read.csv("202208-divvy-tripdata.csv"),
                             read.csv("202209-divvy-tripdata.csv"),
                             read.csv("202210-divvy-tripdata.csv"),
                             read.csv("202211-divvy-tripdata.csv"),
                             read.csv("202212-divvy-tripdata.csv"),
                             read.csv("202301-divvy-tripdata.csv"),
                             read.csv("202302-divvy-tripdata.csv"),
                             read.csv("202303-divvy-tripdata.csv"),
                             read.csv("202304-divvy-tripdata.csv"),
                             read.csv("202305-divvy-tripdata.csv"),
                             read.csv("202306-divvy-tripdata.csv")
)

```


## Clean Data

Data is then cleaned and processed.
This includes:

1. Converting columns:
  a. 'start_time'
  b. 'end_time'
  into date formats.
  
```{r}
cyclistic_year_date$start_time <- ymd_hms(cyclistic_year_date$started_at)
cyclistic_year_date$end_time <- ymd_hms(cyclistic_year_date$ended_at)
```
  
2. Separating 'start_time' column into 'YYY-MM' and 'day' format.

```{r}
cyclistic_year_date$date <- as.Date(cyclistic_year_date$started_at)
cyclistic_year_date$month <- format(as.Date(cyclistic_year_date$date), "%y-%m")
cyclistic_year_date$day <- format(as.Date(cyclistic_year_date$date), "%d")
```

3. Convert 'day' column into weekday name

```{r}
cyclistic_year_date$weekday <- format(as.Date(cyclistic_year_date$date), "%A")
```

4. Calculate the 'ride_length' by using the 'difftime()' function subtracting columns ('ended_at' - 'started_at') into seconds

```{r}
cyclistic_year_date$ride_length <- difftime(cyclistic_year_date$ended_at,
                                            cyclistic_year_date$started_at,
                                            units = "auto")
```

5. Calculate the 'ride_distance' by using the 'distGeo()' function to calculate the distance between start and end locations into km.

```{r}
cyclistic_year_date$ride_distance <- distGeo(
  matrix(c(cyclistic_year_date$start_lng, cyclistic_year_date$start_lat), ncol = 2),
  matrix(c(cyclistic_year_date$end_lng, cyclistic_year_date$end_lat), ncol = 2)
) /1000
```


6. Removing irrelevant data by filtering out and 'ride_distance' over 0 and below 150,
and removing any 'ride_length' values below or equal to 0.

```{r}
cleaned_data <- filter(cyclistic_year_date, ride_distance > 0) %>% 
  filter(ride_distance < 150) %>% 
  filter(ride_length > 0)
```

7. Replace instances of "docked_bike" with "electric_bike" using the 'gsub()' function.

```{r}
cleaned_data$rideable_type <- gsub("docked_bike", "electric_bike", cleaned_data$rideable_type)
```


## Summarize the data

We then take the clean data and group it by member type('member_casual').
We then summarize the grouped data by:
  *average_ride_length
  *max_ride_length
  *median_ride_length
  *min_ride_length
  *average_ride_distance
  *max_ride_distance
  *median_ride_distance
  *min_ride_distance
  
Then convert the average ride length to minutes for readability

```{r cars}
cyc_yr_data <- cleaned_data %>% 
  group_by(member_casual) %>% 
  summarise(amount_of_rides = length(ride_id),
            average_ride_legnth = mean(ride_length),
            max_ride_length = max(ride_length),
            median_ride_length = median(ride_length),
            min_ride_length = min(ride_length),
            avererage_ride_distance = mean(ride_distance),
            max_ride_length = max(ride_distance),
            median_ride_distance = median(ride_distance),
            min_ride_distance = min(ride_distance)
  )

cyc_yr_data$average_ride_legnth <- as.numeric(cyc_yr_data$average_ride_legnth, units = "mins")

```


We then group by member type('memeber_casual') and 'month'

```{r}
cyc_yr_month <- cleaned_data %>% 
  group_by(member_casual, month) %>% 
  summarise(amount_of_rides = length(ride_id),
            average_ride_legnth = mean(ride_length),
            max_ride_length = max(ride_length),
            median_ride_length = median(ride_length),
            min_ride_length = min(ride_length),
            avererage_ride_distance = mean(ride_distance),
            max_ride_length = max(ride_distance),
            median_ride_distance = median(ride_distance),
            min_ride_distance = min(ride_distance)
  )

cyc_yr_month$average_ride_legnth <- as.numeric(cyc_yr_month$average_ride_legnth, units = "mins")
```

We then group by member type('member_casual') and 'weekday'.
Then order the results by days of the week (Monday - Sunday)

```{r}
cyc_yr_wkd <- cleaned_data %>% 
  group_by(member_casual, weekday) %>% 
  summarise(amount_of_rides = length(ride_id),
            average_ride_legnth = mean(ride_length),
            max_ride_length = max(ride_length),
            median_ride_length = median(ride_length),
            min_ride_length = min(ride_length),
            avererage_ride_distance = mean(ride_distance),
            max_ride_length = max(ride_distance),
            median_ride_distance = median(ride_distance),
            min_ride_distance = min(ride_distance)
  )

cyc_yr_wkd$weekday <- factor(cyc_yr_wkd$weekday,
                             levels = c("Monday", "Tuesday", 
                                        "Wednesday", "Thursday", 
                                        "Friday", "Saturday", "Sunday"))

cyc_yr_wkd2 <- cyc_yr_wkd[order(cyc_yr_wkd$weekday), ]

cyc_yr_wkd2$average_ride_legnth <- as.numeric(cyc_yr_wkd2$average_ride_legnth,  units = "mins")
```

We then group by member type('member_casual') and type of ride('ride_type')

```{r}
cyc_yr_type <- cleaned_data %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(amount_of_rides = length(ride_id),
            average_ride_legnth = mean(ride_length),
            max_ride_length = max(ride_length),
            median_ride_length = median(ride_length),
            min_ride_length = min(ride_length),
            avererage_ride_distance = mean(ride_distance),
            max_ride_length = max(ride_distance),
            median_ride_distance = median(ride_distance),
            min_ride_distance = min(ride_distance)
  )
```


## Ploting the data

### Ploting by Month

We visualize the data summarization  by length, count, and distance for each month:

```{r pressure, echo=FALSE}
#Ride length by month

ggplot(cyc_yr_month) +
  geom_col(aes(x = month, y = average_ride_legnth, 
                 fill = member_casual), position = "dodge")+
  labs(title = "Ride Length by Month", 
       subtitle = "Ride Data 22/06 - 23/06",
       x = "Month", 
       y = "Average Ride Length (mins)")

#Ride count by month

ggplot(cyc_yr_month) +
  geom_line(aes(x = month, y = amount_of_rides, 
               color = member_casual, group = member_casual))+
  labs(title = "Number of Rides by Month", 
       subtitle = "Ride Data 22/06 - 23/06",
       x = "Month", 
       y = "Number of Rides")+
  scale_y_continuous(labels = 
                       label_number(scale = 1e-4, suffix ='k', accuracy = 1))

#Ride distance by month

ggplot(cyc_yr_month) +
  geom_line(aes(x = month, y = avererage_ride_distance, 
                color = member_casual, group = member_casual))+
  labs(title = "Disatance of  Rides by Month", 
       subtitle = "Ride Data 22/06 - 23/06",
       x = "Month", 
       y = "Average Distance Length (km)")
```

### Ploting by Weekday

We visualize the data summarization by length, count, and distance for each weekday:

```{r}
#Average length by Weekday

ggplot(cyc_yr_wkd2) +
  geom_col(aes(x = weekday, y = average_ride_legnth, 
               fill = member_casual), position = "dodge")+
  labs(title = "Ride Length by Weekday", 
       subtitle = "Ride Data 22/06 - 23/06",
       x = "Weekday", 
       y = "Average Ride Length (mins)")


#Ride count by Weekday

ggplot(cyc_yr_wkd2) +
  geom_line(aes(x = weekday, y = amount_of_rides, 
               color = member_casual, group = member_casual))+
  labs(title = "Number of Rides by Weekday", 
       subtitle = "22/06 - 23/-06",
       x = "Weekday", 
       y = "Number of Rides") +
  scale_y_continuous(labels = 
                       label_number(scale = 1e-4, suffix ='k', accuracy = 1))

#Average distance by Weekday

ggplot(cyc_yr_wkd2) +
  geom_line(aes(x = weekday, y = avererage_ride_distance, 
               color = member_casual, group = member_casual), position = "dodge")+
  labs(title = "Disatance of  Rides by Weekday", 
       subtitle = "Ride Data 22/06 - 23/06",
       x = "Weekday", 
       y = "Average Distance Length (km)")
```

### Ploting by Weekday

We visualize which ride type is more favorable for each membership:

```{r}
ggplot(cyc_yr_type) +
  geom_col(aes(x = rideable_type, y = amount_of_rides,
               fill = member_casual), position = "dodge") +
  labs(title = "Perferred Type of Ride by Membership",
       subtitle = "Ride Data 22/06 - 23/06",
       x = "Ride Type",
       y = "Number of Rides (Millions)",
       fill = "Membership") +
  scale_y_continuous(
    labels = function(x) format(x/1e6, scientific = FALSE, suffix = "M"))
```

## Findings

The analysis provides valuable insights into the Cyclistic bike share service usage patterns based on various factors, including month, weekdays, and ride types. Key findings include:

* Both member types, Members and Casuals, are more likely to use the Cyclistic service during the warmer months (April - October), with lower usage during the colder months (November - March).
* Members tend to use the service more frequently, while casual users on average spend more time and cover a greater distance per ride compared to members.
* Weekday usage is higher among members, whereas casuals show a preference for using the service during weekends.
* Both user types favor electric bikes, with casual users being significantly more likely to choose electric bikes over classic bikes compared to members.

These insights can inform targeted marketing strategies and service improvements to cater to
different user preferences and optimize the overall bike share experience.

## Proposal

* Distinct usage patterns observed between Casual users and Members of Cyclist bikes
* Precise reasons for differences are inconclusive, requiring a deeper analysis
* Conducting a granular examination to gain comprehensive insights into user behavior
* Aim for a successful and impactful conversion campaign

